---
title: "Practical Machine Learning Course Report"
author: "Pablo Adames"
date: "October 25, 2015"
output: html_document
---

# Summary

Generate a  model  to predict the type of  dumbell biceps curl 
executed by the subjects in the __Weight Lifting Exercises Dataset__
following the classification given by expert trainers. 


# Load data

Data source: [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har)

```{r, load data from disk, cache=TRUE, warning=FALSE}
# read the data for the project as training and test groups
trndat<-read.csv(file="data/pml-training.csv")
tstdat<-read.csv(file="data/pml-testing.csv")
```

# Cleaning data set

A first inspection showed there were too many columns with `NA` values.
A first approach was to get rid of those columns. A total of 36 variables
were discarded. Then the `NA`s were imputed using the `knnImpute`
method and the data was centered and scaled through the `preProcess`
function from the package `caret`.

The number of predictors went from an initial 159 to 123 after imputing and
finally to 57 after eliminating all remaining `NA`s remaining 
with the custom function `extractColumnsWithTooManyNAs` as shown below.

```{r, cleaning data,cache=TRUE,warning=FALSE}
library(caret)
# get the column class
column_class <- function( col ){ifelse(is.data.frame(col), class(col[[1]]), class(col))}
extractNumericCols <- function( df ) {
  # get the column names to index inot the data frame
  l_col_names <- colnames(df)
  # nullify the columns with non-numeric values, return list of data.frames
  l <- sapply(l_col_names, 
              function(x) { cls <- column_class( df[x] );
              ifelse(cls %in% c("numeric","integer"), df[x], "NULL")})
  # filter the columns where there are only numeric values
  isNULL <- which(l[names(l)] =="NULL"); ll <- l[-isNULL] 
  # compose a data frame with the numeric columns (data.frames)
  ndf <- do.call(cbind.data.frame,ll)
}
# new data frame with only numeric columns:
trndf <- extractNumericCols(trndat)
# add outcome column as column #1:
trndf <- cbind(trndat$classe,trndf,deparse.level = 1)
# filter the numeric columns in the test set
numericColNames <- colnames(trndf)
tstdf <- tstdat[,which(colnames(tstdat) %in% numericColNames)]
#Impute and standarize:
preObj <- preProcess(trndf[,-1],method = "knnImpute")

#if there are still columns with NAs eliminate them altogether
extractColumnsWithTooManyNAs <- function( df){
  l <- sapply(colnames(df), function(x) any( is.na(df[x][[1]]) ) )
  df[-which(l)]
}
# eliminate all columns with NAs in them
trnD <- extractColumnsWithTooManyNAs(trndf)
tstD <- extractColumnsWithTooManyNAs(tstdf)
print(paste0(dim(trnD)[1], " training observations", " on ", dim(trnD)[2], " variables after cleaning data."))
```

# Exploratory analysys

Principal component analysis was used due to the high number of predictors.
A five component PCA shows that there are visible clusters for the PC1 and PC2 
interaction. There is only one outlier from this analysis. Further plots of PC1
Vs. PC2 show that only PC5 conditions the output in an important way.

```{r, exploratory analysis with PCA, cache=TRUE,warning=FALSE,fig.align='center'}
preproc <- preProcess(trnD[-1],method = "pca",pcaComp = 5)
dumbbellPC <- predict(preproc,trnD[-1])
featurePlot(x=dumbbellPC,y=trnD$`trndat$classe`,plot="pairs")
```

# Generating a predictive model

The non-linear behaviour observed in the pricipal component analysis
motivated a random forest algorithm to be used for predicting the
type of the dumbbell bicep curl.

```{r, random forest model, cache=TRUE, warning=FALSE}
library(doParallel)
# cl<-makeCluster(detectCores())
cl<-makeCluster(4)
registerDoParallel(cl)
require(randomForest)
dumbbellPC$classe <- trnD$`trndat$classe`
trainD<-createDataPartition(y=dumbbellPC$classe,p=0.3,list = FALSE)
trainingPCsmall <-dumbbellPC[trainD,]
```
```
rf_model_PC <- train(classe~.,data=trainingPCsmall,
                     method="rf",
                     trControl=trainControl(method="cv",number=5),
                     prox=TRUE,
                     allowParallel=TRUE)
rf_model_PC
```
This model has 
Another model can be build using all 53 variables:

```
trnD$classe <- trnD$`trndat$classe`
trnD$`trndat$classe`<- NULL
rf_model_full <- train(classe~.,data=trnD,
                  method="rf",
                  trControl=trainControl(method="cv",number=5),
                  prox=TRUE,
                  allowParallel=TRUE)
rf_model_full
```
